---
title: "SCAMP hom/het validation mineralization analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, message=FALSE, warning=FALSE}
## ================================================================
## SCAMP hom/het validation mineralization analysis
## ---------------------------------------------------------------
## Inputs expected in the working directory by default:
##   1) COMBINED_PIXEL_DATA_SCAMP_output.csv
##      SCAMP pixel-profile matrix, semicolon-delimited.
##      Expected fixed columns: image_name/fish, genotype,
##      followed by pixel columns.
##
## Outputs are written to Results/Figures, Results/Statistics,
## and Results/QC.
## ================================================================

required_packages <- c(
  "tidyverse",
  "ComplexHeatmap",
  "circlize",
  "matrixStats",
  "grid",
  "patchwork",
  "knitr",
  "car"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    "Missing required package(s): ",
    paste(missing_packages, collapse = ", "),
    ". Install them before running this notebook."
  )
}

library(tidyverse)
library(ComplexHeatmap)
library(circlize)
library(matrixStats)
library(grid)
library(patchwork)
library(knitr)
library(car)

## -----------------------------
## User-editable input path
## -----------------------------

scamp_data_path <- "COMBINED_PIXEL_DATA_SCAMP_output.csv"

output_dir <- "Results"
figure_dir <- file.path(output_dir, "Figures")
statistics_dir <- file.path(output_dir, "Statistics")
qc_dir <- file.path(output_dir, "QC")

dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(statistics_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(qc_dir, recursive = TRUE, showWarnings = FALSE)

## -----------------------------
## Plot settings
## -----------------------------

genotype_cols <- c(
  "2_het" = "#261A68",
  "3_mut" = "#B61515"
)

genotype_labels <- c(
  "2_het" = "abcc6a+/-",
  "3_mut" = "abcc6a-/-"
)

scamp_theme <- theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text = element_text(colour = "black"),
    legend.title = element_blank()
  )
```

# 1. Helper functions

```{r helper-functions}
## -----------------------------
## Profile length normalization
## -----------------------------

normalize_row <- function(row, target_length = 1000) {
  row <- as.numeric(row)
  row <- row[is.finite(row)]

  if (length(row) == 0) {
    return(rep(NA_real_, target_length))
  }

  if (length(row) == 1) {
    return(rep(row, target_length))
  }

  approx(
    x = seq_along(row),
    y = row,
    xout = seq(1, length(row), length.out = target_length),
    rule = 2
  )$y
}

normalize_dataframe <- function(df, target_length = 1000, fixed_columns = 2) {
  fixed_data <- df[, 1:fixed_columns, drop = FALSE]
  variable_data <- df[, (fixed_columns + 1):ncol(df), drop = FALSE]

  normalized_data <- variable_data %>%
    apply(1, normalize_row, target_length = target_length) %>%
    t() %>%
    as.data.frame()

  colnames(normalized_data) <- paste0("V", seq_len(target_length))

  cbind(fixed_data, normalized_data)
}

## -----------------------------
## Dataset-wide positive threshold
## -----------------------------

estimate_positive_threshold_low_tail <- function(pixel_matrix,
                                                 background_quantile = 0.60,
                                                 multiplier = 3) {
  values <- as.numeric(pixel_matrix)
  values <- values[is.finite(values)]
  values <- values[values >= 0]

  cutoff <- quantile(values, background_quantile, na.rm = TRUE)
  bg_values <- values[values <= cutoff]

  baseline <- median(bg_values, na.rm = TRUE)
  noise_mad <- mad(bg_values, constant = 1.4826, na.rm = TRUE)

  threshold <- baseline + multiplier * noise_mad

  tibble(
    threshold = threshold,
    baseline = baseline,
    noise_mad = noise_mad,
    background_quantile = background_quantile,
    multiplier = multiplier,
    method = "low-tail median + multiplier*MAD"
  )
}

## -----------------------------
## P-value formatting
## -----------------------------

format_p_stars <- function(p) {
  case_when(
    is.na(p) ~ "n.c.",
    p < 0.0001 ~ "****",
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    TRUE ~ "ns"
  )
}

format_p_value <- function(p) {
  ifelse(is.na(p), NA_character_, formatC(p, format = "e", digits = 3))
}

## -----------------------------
## Robust statistical helpers
## -----------------------------

safe_welch_p <- function(data, value_col, group_col) {
  df <- data %>%
    select(value = all_of(value_col), group = all_of(group_col)) %>%
    filter(!is.na(value), !is.na(group)) %>%
    mutate(group = droplevels(as.factor(group)))

  if (n_distinct(df$group) < 2) return(NA_real_)
  if (any(table(df$group) < 2)) return(NA_real_)
  if (sd(df$value, na.rm = TRUE) < .Machine$double.eps) return(NA_real_)

  tryCatch(
    t.test(value ~ group, data = df, var.equal = FALSE)$p.value,
    error = function(e) NA_real_
  )
}

safe_levene_p <- function(data, value_col, group_col) {
  df <- data %>%
    select(value = all_of(value_col), group = all_of(group_col)) %>%
    filter(!is.na(value), !is.na(group)) %>%
    mutate(group = droplevels(as.factor(group)))

  if (n_distinct(df$group) < 2) return(NA_real_)
  if (any(table(df$group) < 2)) return(NA_real_)
  if (sd(df$value, na.rm = TRUE) < .Machine$double.eps) return(NA_real_)

  tryCatch(
    {
      test <- car::leveneTest(value ~ group, data = df, center = median)
      as.numeric(test[["Pr(>F)"]][1])
    },
    error = function(e) NA_real_
  )
}

safe_one_way_anova <- function(data, value_col, group_col, metric_id, metric_label) {
  df <- data %>%
    select(value = all_of(value_col), group = all_of(group_col)) %>%
    filter(!is.na(value), !is.na(group)) %>%
    mutate(group = droplevels(as.factor(group)))

  if (n_distinct(df$group) < 2) {
    return(tibble(
      metric_id = metric_id,
      metric = metric_label,
      term = NA_character_,
      p_value = NA_real_,
      p_value_formatted = NA_character_,
      significance = "n.c."
    ))
  }

  tryCatch(
    {
      model <- aov(value ~ group, data = df)
      as.data.frame(summary(model)[[1]]) %>%
        tibble::rownames_to_column("term") %>%
        mutate(
          metric_id = metric_id,
          metric = metric_label,
          p_value = `Pr(>F)`,
          p_value_formatted = format_p_value(p_value),
          significance = format_p_stars(p_value)
        ) %>%
        select(metric_id, metric, term, everything())
    },
    error = function(e) {
      tibble(
        metric_id = metric_id,
        metric = metric_label,
        term = NA_character_,
        p_value = NA_real_,
        p_value_formatted = NA_character_,
        significance = "n.c."
      )
    }
  )
}

max_or_one <- function(x) {
  out <- suppressWarnings(max(x, na.rm = TRUE))
  if (!is.finite(out) || out <= 0) out <- 1
  out
}
```

# 2. Load and normalize SCAMP fluorescence profiles

```{r load-scamp-data}
if (!file.exists(scamp_data_path)) {
  stop("SCAMP input file not found: ", scamp_data_path)
}

scamp_raw <- read.csv2(
  scamp_data_path,
  header = TRUE,
  check.names = FALSE
)

scamp_norm <- normalize_dataframe(
  scamp_raw,
  target_length = 1000,
  fixed_columns = 2
)

pixel_data <- scamp_norm %>%
  select(starts_with("V")) %>%
  as.matrix()

storage.mode(pixel_data) <- "numeric"

fish_info <- scamp_norm[[1]]
genotype_info_raw <- as.character(scamp_norm[[2]])

keep_rows <- genotype_info_raw %in% names(genotype_labels)

pixel_data <- pixel_data[keep_rows, , drop = FALSE]
fish_info <- fish_info[keep_rows]
genotype_info_raw <- genotype_info_raw[keep_rows]

genotype_info <- factor(
  genotype_info_raw,
  levels = names(genotype_labels),
  labels = genotype_labels
)

kable(
  tibble(
    n_specimens = nrow(pixel_data),
    n_positions = ncol(pixel_data),
    genotypes = paste(levels(genotype_info), collapse = ", ")
  ),
  caption = "Loaded and normalized SCAMP profile matrix"
)
```

# 3. Build SCAMP biomarker table

```{r metric-table}
threshold_info <- estimate_positive_threshold_low_tail(
  pixel_data,
  background_quantile = 0.60,
  multiplier = 3
)

positive_threshold <- threshold_info$threshold[[1]]

write.csv2(
  threshold_info,
  file.path(qc_dir, "SCAMP_positive_threshold_info.csv"),
  row.names = FALSE
)

scamp_metric_df <- data.frame(
  fish = fish_info,
  genotype_raw = genotype_info_raw,

  integrated_fluorescence = rowSums(
    pixel_data,
    na.rm = TRUE
  ),

  positive_area = apply(
    pixel_data,
    1,
    function(x) mean(x > positive_threshold, na.rm = TRUE) * 100
  ),

  mean_positive_fluorescence = apply(
    pixel_data,
    1,
    function(x) {
      positive_pixels <- x[x > positive_threshold]
      if (length(positive_pixels) == 0) return(NA_real_)
      mean(positive_pixels, na.rm = TRUE)
    }
  )
) %>%
  mutate(
    genotype = factor(
      genotype_raw,
      levels = names(genotype_labels),
      labels = genotype_labels
    )
  ) %>%
  filter(!is.na(genotype))

write.csv2(
  scamp_metric_df,
  file.path(statistics_dir, "SCAMP_hom_het_fluorescence_metrics_per_fish.csv"),
  row.names = FALSE
)

kable(threshold_info, caption = "Dataset-wide positive-threshold estimate")
kable(head(scamp_metric_df), caption = "SCAMP-derived fluorescence metrics")
```

# 4. Unified two-genotype analysis and plotting function

```{r two-group-analysis-function}
make_two_group_metric_analysis <- function(data,
                                           metric,
                                           y_label,
                                           plot_title,
                                           metric_id = NULL) {
  metric <- rlang::ensym(metric)
  metric_name <- rlang::as_string(metric)
  if (is.null(metric_id)) metric_id <- metric_name

  plot_df <- data %>%
    filter(!is.na(.data[[metric_name]]))

  ## One-way ANOVA is exported for completeness.
  ## For two groups, Welch's t-test is used as the primary plotted comparison.
  anova_table <- safe_one_way_anova(
    plot_df,
    value_col = metric_name,
    group_col = "genotype",
    metric_id = metric_id,
    metric_label = plot_title
  )

  p_welch <- safe_welch_p(
    plot_df,
    value_col = metric_name,
    group_col = "genotype"
  )

  welch_table <- tibble(
    metric_id = metric_id,
    metric = plot_title,
    comparison = "abcc6a+/- vs abcc6a-/-",
    test = "Welch two-sample t-test",
    p_raw = p_welch,
    p_adj_BH = p_welch,
    p_label = format_p_stars(p_welch),
    p_raw_formatted = format_p_value(p_welch),
    p_adj_BH_formatted = format_p_value(p_welch)
  )

  y_pos <- max_or_one(plot_df[[metric_name]]) * 1.12

  sig_df <- tibble(
    x = 1,
    xend = 2,
    y = y_pos,
    yend = y_pos,
    label_x = 1.5,
    label_y = y_pos * 1.04,
    label = format_p_stars(p_welch)
  )

  plot <- ggplot(
    plot_df,
    aes(
      x = genotype,
      y = .data[[metric_name]],
      fill = genotype,
      colour = genotype
    )
  ) +
    geom_boxplot(
      width = 0.22,
      linewidth = 0.8,
      alpha = 0.45,
      outlier.shape = NA
    ) +
    geom_jitter(
      width = 0.12,
      size = 2.5,
      alpha = 0.85
    ) +
    geom_segment(
      data = sig_df,
      aes(x = x, xend = xend, y = y, yend = yend),
      inherit.aes = FALSE,
      linewidth = 0.5,
      colour = "black"
    ) +
    geom_text(
      data = sig_df,
      aes(x = label_x, y = label_y, label = label),
      inherit.aes = FALSE,
      size = 5
    ) +
    scale_fill_manual(values = setNames(unname(genotype_cols), genotype_labels)) +
    scale_colour_manual(values = setNames(unname(genotype_cols), genotype_labels)) +
    coord_cartesian(ylim = c(0, y_pos * 1.18), clip = "off") +
    scamp_theme +
    theme(legend.position = "none") +
    labs(
      x = "Genotype",
      y = y_label,
      title = plot_title
    )

  list(
    plot = plot,
    anova = anova_table,
    welch = welch_table,
    data = plot_df
  )
}
```

# 5. SCAMP fluorescence metric analyses

```{r fluorescence-metric-analyses}
integrated_result <- make_two_group_metric_analysis(
  data = scamp_metric_df,
  metric = integrated_fluorescence,
  y_label = "Integrated fluorescence (a.u.)",
  plot_title = "Total mineralization burden",
  metric_id = "integrated_fluorescence"
)

positive_area_result <- make_two_group_metric_analysis(
  data = scamp_metric_df,
  metric = positive_area,
  y_label = paste0(
    "Positive area (%)\nThreshold = ",
    round(positive_threshold, 2),
    " a.u."
  ),
  plot_title = "Extent of spinal mineralization",
  metric_id = "positive_area"
)

mean_positive_result <- make_two_group_metric_analysis(
  data = scamp_metric_df,
  metric = mean_positive_fluorescence,
  y_label = paste0(
    "Mean positive fluorescence (a.u.)\nThreshold = ",
    round(positive_threshold, 2),
    " a.u."
  ),
  plot_title = "Mineralization density",
  metric_id = "mean_positive_fluorescence"
)

p_integrated <- integrated_result$plot
p_positive_area <- positive_area_result$plot
p_density <- mean_positive_result$plot

fluorescence_anova_results <- bind_rows(
  integrated_result$anova,
  positive_area_result$anova,
  mean_positive_result$anova
)

fluorescence_welch_results <- bind_rows(
  integrated_result$welch,
  positive_area_result$welch,
  mean_positive_result$welch
)

write.csv2(
  fluorescence_anova_results,
  file.path(statistics_dir, "SCAMP_hom_het_one_way_ANOVA_results.csv"),
  row.names = FALSE
)

write.csv2(
  fluorescence_welch_results,
  file.path(statistics_dir, "SCAMP_hom_het_Welch_results.csv"),
  row.names = FALSE
)

kable(fluorescence_anova_results, caption = "SCAMP fluorescence metric one-way ANOVA results")
kable(fluorescence_welch_results, caption = "SCAMP fluorescence metric Welch test results")
```

# 6. Integrated fluorescence variability analysis

```{r integrated-variance-analysis}
## Brown-Forsythe Levene test: genotype-specific variance comparison
## for integrated fluorescence. This follows the same logic as the
## MB-PPi analysis, but here there is only one genotype comparison,
## so BH-adjusted p is identical to the raw p value.

integrated_variance_summary <- scamp_metric_df %>%
  filter(!is.na(integrated_fluorescence), !is.na(genotype)) %>%
  group_by(genotype) %>%
  summarise(
    n = n(),
    mean = mean(integrated_fluorescence, na.rm = TRUE),
    median = median(integrated_fluorescence, na.rm = TRUE),
    sd = sd(integrated_fluorescence, na.rm = TRUE),
    variance = var(integrated_fluorescence, na.rm = TRUE),
    cv_percent = ifelse(mean > 0, sd / mean * 100, NA_real_),
    .groups = "drop"
  )

integrated_levene_results <- tibble(
  metric_id = "integrated_fluorescence",
  metric = "Integrated fluorescence",
  comparison = "abcc6a+/- vs abcc6a-/-",
  test = "Brown-Forsythe Levene test; center = median",
  p_raw = safe_levene_p(
    scamp_metric_df,
    value_col = "integrated_fluorescence",
    group_col = "genotype"
  )
) %>%
  mutate(
    p_adj_BH = p_raw,
    p_label = format_p_stars(p_adj_BH),
    p_raw_formatted = format_p_value(p_raw),
    p_adj_BH_formatted = format_p_value(p_adj_BH)
  )

write.csv2(
  integrated_variance_summary,
  file.path(statistics_dir, "SCAMP_hom_het_integrated_fluorescence_variance_summary.csv"),
  row.names = FALSE
)

write.csv2(
  integrated_levene_results,
  file.path(statistics_dir, "SCAMP_hom_het_integrated_fluorescence_Levene_BrownForsythe_results.csv"),
  row.names = FALSE
)

kable(integrated_variance_summary, caption = "Integrated fluorescence variance summary by genotype")
kable(integrated_levene_results, caption = "Integrated fluorescence variance comparison between genotypes")
```

# 7. Heatmap visualization

```{r heatmap-original-style, fig.width=12, fig.height=8}
## The heatmap is exported separately because ComplexHeatmap is not a ggplot object.
## Rows represent specimens; columns represent normalized spine positions.
## Color shows row-wise relative pattern intensity, while the barplot annotation
## shows integrated fluorescence per specimen.

total_intensity <- rowSums(pixel_data, na.rm = TRUE)

row_max <- rowMaxs(as.matrix(pixel_data), na.rm = TRUE)
row_max[!is.finite(row_max) | row_max <= 0] <- 1

pixel_pattern <- sweep(
  pixel_data,
  1,
  row_max,
  "/"
)

pixel_pattern[is.na(pixel_pattern)] <- 0

col_fun_pattern <- colorRamp2(
  c(0, 0.5, 1),
  c("blue", "white", "red")
)

ha <- rowAnnotation(
  Genotype = genotype_info,
  `Integrated fluorescence\n(a.u.)` = anno_barplot(
    total_intensity,
    border = FALSE,
    gp = gpar(fill = "grey40")
  ),
  col = list(
    Genotype = setNames(unname(genotype_cols), genotype_labels)
  )
)

heatmap <- Heatmap(
  pixel_pattern,
  name = "Pattern\n(relative intensity)",
  col = col_fun_pattern,
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  right_annotation = ha,
  heatmap_legend_param = list(
    title = "Pattern\nrelative intensity",
    at = c(0, 0.5, 1),
    labels = c("0", "0.5", "1")
  )
)

pdf(
  file.path(figure_dir, "SCAMP_hom_het_heatmap.pdf"),
  width = 12,
  height = 8
)
draw(heatmap)
dev.off()

png(
  file.path(figure_dir, "SCAMP_hom_het_heatmap.png"),
  width = 2400,
  height = 1600,
  res = 200
)
draw(heatmap)
dev.off()

heatmap
```

# 8. Ribbon plot of spatial mineralization pattern

```{r ribbon-plot, fig.width=12, fig.height=5}
long_df <- scamp_metric_df %>%
  select(fish, genotype) %>%
  bind_cols(as.data.frame(pixel_data)) %>%
  pivot_longer(
    cols = starts_with("V"),
    names_to = "Pixel",
    values_to = "Intensity"
  ) %>%
  mutate(
    Pixel = as.numeric(gsub("V", "", Pixel)),
    Intensity = as.numeric(Intensity)
  )

ribbon_df <- long_df %>%
  group_by(Pixel, genotype) %>%
  summarise(
    n = sum(!is.na(Intensity)),
    mean = mean(Intensity, na.rm = TRUE),
    median = median(Intensity, na.rm = TRUE),
    sd = sd(Intensity, na.rm = TRUE),
    sem = sd / sqrt(n),
    CI_lower = ifelse(
      n > 1,
      pmax(mean - qt(0.975, df = n - 1) * sem, 0),
      NA_real_
    ),
    CI_upper = ifelse(
      n > 1,
      mean + qt(0.975, df = n - 1) * sem,
      NA_real_
    ),
    .groups = "drop"
  )

positive_regions <- ribbon_df %>%
  group_by(genotype) %>%
  arrange(Pixel, .by_group = TRUE) %>%
  mutate(
    positive = mean > positive_threshold,
    region_id = cumsum(positive != lag(positive, default = FALSE))
  ) %>%
  filter(positive) %>%
  group_by(genotype, region_id) %>%
  summarise(
    xmin = min(Pixel, na.rm = TRUE),
    xmax = max(Pixel, na.rm = TRUE),
    .groups = "drop"
  )

p_ribbon <- ggplot(
  ribbon_df,
  aes(
    x = Pixel,
    y = mean,
    colour = genotype,
    fill = genotype
  )
) +
  geom_rect(
    data = positive_regions,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = -Inf,
      ymax = Inf,
      fill = genotype
    ),
    inherit.aes = FALSE,
    alpha = 0.10,
    colour = NA
  ) +
  geom_ribbon(
    aes(ymin = CI_lower, ymax = CI_upper),
    colour = NA,
    alpha = 0.35
  ) +
  geom_line(linewidth = 1) +
  geom_hline(
    yintercept = positive_threshold,
    linetype = "dashed",
    colour = "grey35",
    linewidth = 0.4
  ) +
  scale_fill_manual(values = setNames(unname(genotype_cols), genotype_labels)) +
  scale_colour_manual(values = setNames(unname(genotype_cols), genotype_labels)) +
  coord_cartesian(ylim = c(0, max_or_one(ribbon_df$CI_upper) * 1.05)) +
  scamp_theme +
  labs(
    x = "Normalised spine position",
    y = "Mean fluorescence intensity (a.u.)",
    title = "Spatial mineralization profile"
  )

p_ribbon
```

# 9. Biomarker plots

```{r biomarker-plots, fig.width=12, fig.height=4}
p_integrated | p_positive_area | p_density
```

# 10. Final ggplot figure export

```{r final-ggplot-export, fig.width=12, fig.height=8}
## Heatmap is saved separately above. This combined figure contains ggplot panels only.

final_figure <-
  p_ribbon /
  (p_integrated | p_positive_area | p_density) +
  plot_layout(heights = c(1.2, 1)) +
  plot_annotation(
    title = "SCAMP hom/het mineralization analysis",
    tag_levels = "A"
  )

ggsave(
  filename = file.path(figure_dir, "SCAMP_hom_het_summary_ggplots.pdf"),
  plot = final_figure,
  width = 12,
  height = 8,
  units = "in",
  device = cairo_pdf
)

ggsave(
  filename = file.path(figure_dir, "SCAMP_hom_het_summary_ggplots.png"),
  plot = final_figure,
  width = 12,
  height = 8,
  units = "in",
  dpi = 600
)

final_figure
```

# 11. Session information

```{r session-info}
writeLines(
  capture.output(sessionInfo()),
  con = file.path(qc_dir, "sessionInfo.txt")
)

sessionInfo()
```
