---
title: "Human RESQ fused-region analysis"
author: "SCAMP / RESQ analysis workflow"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
## ==========================================================
## Human RESQ fused-region analysis
## ==========================================================
## Purpose:
##   Analyse whether fused-region size is associated with AP-axis position
##   and whether this relationship differs between genotypes.
##
## Expected input columns:
##   fishID, genotype, Area number, Area, AP_axis
##
## Main outputs:
##   Results/Figures/
##   Results/Statistics/
##   Results/QC/
## ==========================================================

suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(cocor)
  library(broom)
})

## -----------------------------
## User settings
## -----------------------------

input_file <- "fusion-human-resq.xlsx"
input_sheet <- 1

results_dir <- "Results"
fig_dir <- file.path(results_dir, "Figures")
stats_dir <- file.path(results_dir, "Statistics")
qc_dir <- file.path(results_dir, "QC")

dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(stats_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(qc_dir, recursive = TRUE, showWarnings = FALSE)

genotype_levels <- c("2_het", "3_mut")

genotype_labels <- c(
  "2_het" = "abcc6a+/-",
  "3_mut" = "abcc6a-/-"
)

genotype_cols <- c(
  "2_het" = "#261A68",
  "3_mut" = "#B61515"
)

## Binning setting for optional AP-axis summary
bin_width <- 100

## -----------------------------
## Helper functions
## -----------------------------

format_p_stars <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "n.c.",
    p < 0.0001 ~ "****",
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    TRUE ~ "ns"
  )
}

safe_cor_test <- function(data, x_col, y_col, method = "kendall") {
  x <- data[[x_col]]
  y <- data[[y_col]]
  ok <- is.finite(x) & is.finite(y)

  if (sum(ok) < 3) {
    return(tibble(
      estimate = NA_real_,
      statistic = NA_real_,
      p.value = NA_real_,
      method = method,
      n = sum(ok)
    ))
  }

  out <- tryCatch(
    cor.test(x[ok], y[ok], method = method, exact = FALSE),
    error = function(e) NULL
  )

  if (is.null(out)) {
    return(tibble(
      estimate = NA_real_,
      statistic = NA_real_,
      p.value = NA_real_,
      method = method,
      n = sum(ok)
    ))
  }

  broom::tidy(out) %>%
    mutate(
      method = method,
      n = sum(ok)
    )
}

write_session_info <- function(path) {
  sink(path)
  print(sessionInfo())
  sink()
}
```

# 1. Load and clean data

```{r load-data}
raw_dat <- read_excel(input_file, sheet = input_sheet)

required_cols <- c("fishID", "genotype", "Area number", "Area", "AP_axis")
missing_cols <- setdiff(required_cols, names(raw_dat))

if (length(missing_cols) > 0) {
  stop(
    "Missing required column(s): ",
    paste(missing_cols, collapse = ", ")
  )
}

resq_df <- raw_dat %>%
  as_tibble() %>%
  mutate(
    fishID = factor(fishID),
    genotype_raw = as.character(genotype),
    genotype = factor(genotype_raw, levels = genotype_levels),
    genotype_label = factor(
      genotype_labels[as.character(genotype)],
      levels = unname(genotype_labels)
    ),
    `Area number` = factor(`Area number`),
    Area = as.numeric(Area),
    AP_axis = as.numeric(AP_axis)
  ) %>%
  filter(
    !is.na(genotype),
    is.finite(Area),
    is.finite(AP_axis)
  )

if (nrow(resq_df) == 0) {
  stop("No usable rows remained after filtering. Check genotype labels and numeric columns.")
}

write_csv(
  resq_df,
  file.path(stats_dir, "human_RESQ_cleaned_input_data.csv")
)

resq_df
```

# 2. Sample overview

```{r sample-overview}
sample_summary <- resq_df %>%
  group_by(genotype, genotype_label) %>%
  summarise(
    n_regions = n(),
    n_fish = n_distinct(fishID),
    mean_area = mean(Area, na.rm = TRUE),
    median_area = median(Area, na.rm = TRUE),
    sd_area = sd(Area, na.rm = TRUE),
    min_area = min(Area, na.rm = TRUE),
    max_area = max(Area, na.rm = TRUE),
    mean_AP_axis = mean(AP_axis, na.rm = TRUE),
    min_AP_axis = min(AP_axis, na.rm = TRUE),
    max_AP_axis = max(AP_axis, na.rm = TRUE),
    .groups = "drop"
  )

write_csv(
  sample_summary,
  file.path(stats_dir, "human_RESQ_sample_summary.csv")
)

sample_summary
```

# 3. Kendall correlation analysis

This analysis tests whether fused-region area is associated with AP-axis position. Kendall's rank correlation is used because it is robust for monotonic relationships and does not assume normally distributed variables.

```{r kendall-correlations}
cor_overall <- safe_cor_test(
  resq_df,
  x_col = "Area",
  y_col = "AP_axis",
  method = "kendall"
) %>%
  mutate(group = "all")

cor_by_genotype <- resq_df %>%
  group_by(genotype, genotype_label) %>%
  group_modify(~ safe_cor_test(.x, "Area", "AP_axis", method = "kendall")) %>%
  ungroup() %>%
  mutate(group = as.character(genotype))

cor_results <- bind_rows(
  cor_overall,
  cor_by_genotype
) %>%
  mutate(
    p_label = format_p_stars(p.value)
  )

write_csv(
  cor_results,
  file.path(stats_dir, "human_RESQ_Kendall_correlation_results.csv")
)

cor_results
```

# 4. Genotype-specific correlation comparison

This analysis asks whether the Area-AP relationship differs between genotypes. The `cocor` package is used to compare two independent correlations.

```{r compare-correlations}
resq_het <- resq_df %>% filter(genotype == "2_het")
resq_mut <- resq_df %>% filter(genotype == "3_mut")

dat_list <- list(
  het = as.data.frame(resq_het),
  mut = as.data.frame(resq_mut)
)

cocor_result <- NULL
cocor_text_path <- file.path(stats_dir, "human_RESQ_cocor_correlation_comparison.txt")

if (nrow(resq_het) >= 3 && nrow(resq_mut) >= 3) {
  cocor_result <- cocor(
    formula = ~ Area + AP_axis | Area + AP_axis,
    data = dat_list,
    test = "fisher1925"
  )

  capture.output(
    cocor_result,
    file = cocor_text_path
  )
} else {
  writeLines(
    "cocor comparison was not run because at least one genotype had fewer than 3 observations.",
    cocor_text_path
  )
}

cocor_result
```

# 5. Linear interaction model

This model tests whether the relationship between fused-region area and AP-axis position differs by genotype.

```{r linear-interaction-model}
lm_fit <- lm(
  Area ~ genotype * AP_axis,
  data = resq_df
)

lm_coefficients <- broom::tidy(lm_fit) %>%
  mutate(p_label = format_p_stars(p.value))

lm_anova <- broom::tidy(anova(lm_fit)) %>%
  mutate(p_label = format_p_stars(p.value))

lm_glance <- broom::glance(lm_fit)

write_csv(
  lm_coefficients,
  file.path(stats_dir, "human_RESQ_linear_model_coefficients.csv")
)

write_csv(
  lm_anova,
  file.path(stats_dir, "human_RESQ_linear_model_ANOVA.csv")
)

write_csv(
  lm_glance,
  file.path(stats_dir, "human_RESQ_linear_model_summary.csv")
)

lm_coefficients
lm_anova
lm_glance
```

# 6. Scatterplot with LOESS trend

```{r scatterplot-loess, fig.width=10, fig.height=4}
p_scatter_loess <- ggplot(
  resq_df,
  aes(
    x = AP_axis,
    y = Area,
    colour = genotype,
    fill = genotype
  )
) +
  geom_point(
    alpha = 0.65,
    size = 2
  ) +
  geom_smooth(
    method = "loess",
    se = TRUE,
    alpha = 0.20,
    linewidth = 1
  ) +
  scale_colour_manual(
    values = genotype_cols,
    labels = genotype_labels
  ) +
  scale_fill_manual(
    values = genotype_cols,
    labels = genotype_labels
  ) +
  theme_classic(base_size = 12) +
  labs(
    x = "AP-axis position",
    y = "Fused region area",
    title = "Fused region size along the AP axis",
    colour = "Genotype",
    fill = "Genotype"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text = element_text(colour = "black")
  )

p_scatter_loess

ggsave(
  filename = file.path(fig_dir, "human_RESQ_scatterplot_LOESS.pdf"),
  plot = p_scatter_loess,
  width = 10,
  height = 4,
  device = cairo_pdf
)

ggsave(
  filename = file.path(fig_dir, "human_RESQ_scatterplot_LOESS.png"),
  plot = p_scatter_loess,
  width = 10,
  height = 4,
  dpi = 600
)
```

# 7. Binned AP-axis summary

Because AP-axis position is near-continuous, exact-position groups often contain only one observation. For mean ± SE visualization, AP-axis position is therefore binned.

```{r binned-summary}
ap_min <- floor(min(resq_df$AP_axis, na.rm = TRUE) / bin_width) * bin_width
ap_max <- ceiling(max(resq_df$AP_axis, na.rm = TRUE) / bin_width) * bin_width + bin_width

resq_binned <- resq_df %>%
  mutate(
    AP_bin = cut(
      AP_axis,
      breaks = seq(ap_min, ap_max, by = bin_width),
      include.lowest = TRUE,
      right = FALSE
    ),
    AP_bin_mid = as.numeric(sub("\\[(.*),(.*)\\)", "\\1", AP_bin)) + bin_width / 2
  )

summary_binned <- resq_binned %>%
  group_by(genotype, genotype_label, AP_bin, AP_bin_mid) %>%
  summarise(
    n = n(),
    n_fish = n_distinct(fishID),
    mean_area = mean(Area, na.rm = TRUE),
    sd_area = if_else(n > 1, sd(Area, na.rm = TRUE), NA_real_),
    se_area = if_else(n > 1, sd_area / sqrt(n), NA_real_),
    ymin = mean_area - se_area,
    ymax = mean_area + se_area,
    .groups = "drop"
  )

write_csv(
  summary_binned,
  file.path(stats_dir, "human_RESQ_binned_AP_axis_summary.csv")
)

summary_binned
```

```{r binned-plot, fig.width=10, fig.height=4}
p_binned <- ggplot() +
  geom_point(
    data = resq_df,
    aes(
      x = AP_axis,
      y = Area,
      colour = genotype
    ),
    alpha = 0.35,
    size = 1.8
  ) +
  geom_line(
    data = summary_binned,
    aes(
      x = AP_bin_mid,
      y = mean_area,
      colour = genotype,
      group = genotype
    ),
    linewidth = 1
  ) +
  geom_point(
    data = summary_binned,
    aes(
      x = AP_bin_mid,
      y = mean_area,
      colour = genotype
    ),
    size = 3
  ) +
  geom_errorbar(
    data = summary_binned %>% filter(!is.na(se_area)),
    aes(
      x = AP_bin_mid,
      ymin = ymin,
      ymax = ymax,
      colour = genotype
    ),
    width = bin_width * 0.15,
    linewidth = 0.8
  ) +
  scale_colour_manual(
    values = genotype_cols,
    labels = genotype_labels
  ) +
  theme_classic(base_size = 12) +
  labs(
    x = paste0("AP-axis position, binned every ", bin_width, " units"),
    y = "Fused region area",
    title = "Binned fused-region size along the AP axis",
    colour = "Genotype"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text = element_text(colour = "black")
  )

p_binned

ggsave(
  filename = file.path(fig_dir, "human_RESQ_binned_mean_SE.pdf"),
  plot = p_binned,
  width = 10,
  height = 4,
  device = cairo_pdf
)

ggsave(
  filename = file.path(fig_dir, "human_RESQ_binned_mean_SE.png"),
  plot = p_binned,
  width = 10,
  height = 4,
  dpi = 600
)
```

# 8. Save session information

```{r session-info}
write_session_info(
  file.path(qc_dir, "human_RESQ_sessionInfo.txt")
)
```
