---
title: "SCAMP MB/PPi mineralization analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, message=FALSE, warning=FALSE}
## ================================================================
## SCAMP MB/PPi mineralization analysis
## ---------------------------------------------------------------
## Inputs expected in the working directory by default:
##   1) MB_PPi_data_combined.csv
##      SCAMP pixel-profile matrix, semicolon-delimited.
##      Expected fixed columns: fish/sample, treatment, replicate/other, genotype,
##      followed by pixel columns.
##
##   2) superplota_method_allcalc_MB_PPi.csv
##      Manual calcified-vertebrae quantification, semicolon-delimited.
##      Expected columns: Genotype, Treatment, Calcified.vertebrae.all
##
## Outputs are written to Results/Figures and Results/Statistics.
## ================================================================

required_packages <- c(
  "tidyverse",
  "patchwork",
  "knitr",
  "car"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    "Missing required package(s): ",
    paste(missing_packages, collapse = ", "),
    ". Install them before running this notebook."
  )
}

library(tidyverse)
library(patchwork)
library(knitr)
library(car)

## -----------------------------
## User-editable input paths
## -----------------------------

scamp_data_path <- "MB_PPi_data_combined.csv"
calcified_data_path <- "superplota_method_allcalc_MB_PPi.csv"

output_dir <- "Results"
figure_dir <- file.path(output_dir, "Figures")
statistics_dir <- file.path(output_dir, "Statistics")
qc_dir <- file.path(output_dir, "QC")

dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(statistics_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(qc_dir, recursive = TRUE, showWarnings = FALSE)

## -----------------------------
## Plot settings
## -----------------------------

genotype_cols <- c(
  "2_het" = "#261A68",
  "3_mut" = "#B61515"
)

genotype_labels <- c(
  "2_het" = "abcc6a+/-",
  "3_mut" = "abcc6a-/-"
)

treatment_levels <- c(
  "_-PPi_-MB",
  "_-PPi_+MB",
  "_+PPi_-MB",
  "_+PPi_+MB"
)

treatment_labels <- c(
  "-MB\n-PPi",
  "+MB\n-PPi",
  "-MB\n+PPi",
  "+MB\n+PPi"
)

scamp_theme <- theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text = element_text(colour = "black"),
    legend.title = element_blank()
  )
```

# 1. Helper functions

```{r helper-functions}
## -----------------------------
## Profile length normalization
## -----------------------------

normalize_row <- function(row, target_length = 1000) {
  row <- as.numeric(row)
  row <- row[is.finite(row)]

  if (length(row) == 0) {
    return(rep(NA_real_, target_length))
  }

  if (length(row) == 1) {
    return(rep(row, target_length))
  }

  approx(
    x = seq_along(row),
    y = row,
    xout = seq(1, length(row), length.out = target_length),
    rule = 2
  )$y
}

normalize_dataframe <- function(df, target_length = 1000, fixed_columns = 4) {
  fixed_data <- df[, 1:fixed_columns, drop = FALSE]
  variable_data <- df[, (fixed_columns + 1):ncol(df), drop = FALSE]

  normalized_data <- variable_data %>%
    apply(1, normalize_row, target_length = target_length) %>%
    t() %>%
    as.data.frame()

  colnames(normalized_data) <- paste0("V", seq_len(target_length))

  cbind(fixed_data, normalized_data)
}

## -----------------------------
## Dataset-wide positive threshold
## -----------------------------

estimate_positive_threshold_low_tail <- function(pixel_matrix,
                                                 background_quantile = 0.60,
                                                 multiplier = 3) {
  values <- as.numeric(pixel_matrix)
  values <- values[is.finite(values)]
  values <- values[values >= 0]

  cutoff <- quantile(values, background_quantile, na.rm = TRUE)
  bg_values <- values[values <= cutoff]

  baseline <- median(bg_values, na.rm = TRUE)
  noise_mad <- mad(bg_values, constant = 1.4826, na.rm = TRUE)

  threshold <- baseline + multiplier * noise_mad

  tibble(
    threshold = threshold,
    baseline = baseline,
    noise_mad = noise_mad,
    background_quantile = background_quantile,
    multiplier = multiplier,
    method = "low-tail median + multiplier*MAD"
  )
}

## -----------------------------
## P-value formatting
## -----------------------------

format_p_stars <- function(p) {
  case_when(
    is.na(p) ~ "n.c.",
    p < 0.0001 ~ "****",
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    TRUE ~ "ns"
  )
}

format_p_value <- function(p) {
  ifelse(is.na(p), NA_character_, formatC(p, format = "e", digits = 3))
}

## -----------------------------
## Robust statistical helpers
## -----------------------------

safe_welch_p <- function(data, value_col, group_col) {
  x <- data[[value_col]]
  g <- data[[group_col]]

  if (length(unique(g[!is.na(g)])) < 2) return(NA_real_)

  split_x <- split(x, g)

  if (any(sapply(split_x, function(v) sum(!is.na(v)) < 2))) return(NA_real_)
  if (sd(x, na.rm = TRUE) < .Machine$double.eps) return(NA_real_)

  tryCatch(
    t.test(x ~ g, var.equal = FALSE)$p.value,
    error = function(e) NA_real_
  )
}

safe_levene_p <- function(data, value_col, group_col) {
  df <- data %>%
    select(value = all_of(value_col), group = all_of(group_col)) %>%
    filter(!is.na(value), !is.na(group)) %>%
    mutate(group = droplevels(as.factor(group)))

  if (n_distinct(df$group) < 2) return(NA_real_)
  if (any(table(df$group) < 2)) return(NA_real_)
  if (sd(df$value, na.rm = TRUE) < .Machine$double.eps) return(NA_real_)

  tryCatch(
    {
      test <- car::leveneTest(value ~ group, data = df, center = median)
      as.numeric(test[["Pr(>F)"]][1])
    },
    error = function(e) NA_real_
  )
}
```

# 2. Load and normalize SCAMP fluorescence profiles

```{r load-scamp-data}
if (!file.exists(scamp_data_path)) {
  stop("SCAMP input file not found: ", scamp_data_path)
}

scamp_raw <- read.csv2(scamp_data_path, header = TRUE, check.names = FALSE)

scamp_norm <- normalize_dataframe(
  scamp_raw,
  target_length = 1000,
  fixed_columns = 4
)

pixel_data <- as.matrix(scamp_norm[, 5:ncol(scamp_norm)])
storage.mode(pixel_data) <- "numeric"

fish_info <- scamp_norm[[1]]
treatment_info <- scamp_norm[[2]]
genotype_info <- scamp_norm[[4]]

threshold_info <- estimate_positive_threshold_low_tail(
  pixel_data,
  background_quantile = 0.60,
  multiplier = 3
)

positive_threshold <- threshold_info$threshold[[1]]

write.csv2(
  threshold_info,
  file.path(qc_dir, "SCAMP_positive_threshold_info.csv"),
  row.names = FALSE
)

threshold_info
```

# 3. Build SCAMP biomarker table

```{r metric-table}
scamp_metric_df <- data.frame(
  fish = fish_info,
  treatment_raw = treatment_info,
  genotype_raw = genotype_info,

  integrated_fluorescence = rowSums(
    pixel_data,
    na.rm = TRUE
  ),

  positive_area = apply(
    pixel_data,
    1,
    function(x) mean(x > positive_threshold, na.rm = TRUE) * 100
  ),

  mean_positive_fluorescence = apply(
    pixel_data,
    1,
    function(x) {
      positive_pixels <- x[x > positive_threshold]
      if (length(positive_pixels) == 0) return(NA_real_)
      mean(positive_pixels, na.rm = TRUE)
    }
  )
) %>%
  mutate(
    genotype = factor(
      genotype_raw,
      levels = names(genotype_labels),
      labels = genotype_labels
    ),
    treatment = factor(
      treatment_raw,
      levels = treatment_levels,
      labels = treatment_labels
    )
  ) %>%
  filter(!is.na(genotype), !is.na(treatment))

write.csv2(
  scamp_metric_df,
  file.path(statistics_dir, "SCAMP_fluorescence_metrics_per_fish.csv"),
  row.names = FALSE
)

kable(head(scamp_metric_df), caption = "SCAMP-derived fluorescence metrics")
```

# 4. Unified grouped analysis and plotting function

```{r grouped-analysis-function}
make_grouped_metric_analysis <- function(data,
                                         metric,
                                         y_label,
                                         plot_title,
                                         metric_id = NULL) {
  metric <- rlang::ensym(metric)
  metric_name <- rlang::as_string(metric)
  if (is.null(metric_id)) metric_id <- metric_name

  plot_df <- data %>%
    filter(!is.na(.data[[metric_name]]))

  ## Two-way ANOVA: genotype, treatment, and genotype:treatment interaction
  anova_model <- aov(
    as.formula(paste0(metric_name, " ~ genotype * treatment")),
    data = plot_df
  )

  anova_table <- as.data.frame(summary(anova_model)[[1]]) %>%
    tibble::rownames_to_column("term") %>%
    mutate(
      metric_id = metric_id,
      metric = plot_title,
      p_value = `Pr(>F)`,
      p_value_formatted = format_p_value(p_value),
      significance = format_p_stars(p_value)
    ) %>%
    select(metric_id, metric, term, everything())

  ## Genotype comparisons within each treatment: Welch t-test + BH correction
  welch_table <- plot_df %>%
    group_by(treatment) %>%
    summarise(
      p_raw = safe_welch_p(
        cur_data(),
        value_col = metric_name,
        group_col = "genotype"
      ),
      .groups = "drop"
    ) %>%
    mutate(
      metric_id = metric_id,
      metric = plot_title,
      p_adj_BH = p.adjust(p_raw, method = "BH"),
      p_label = format_p_stars(p_adj_BH),
      p_raw_formatted = format_p_value(p_raw),
      p_adj_BH_formatted = format_p_value(p_adj_BH),
      treatment_x = as.numeric(treatment),
      x = treatment_x - 0.18,
      xend = treatment_x + 0.18
    ) %>%
    select(metric_id, metric, treatment, p_raw, p_adj_BH, p_label, everything())

  y_positions <- plot_df %>%
    group_by(treatment) %>%
    summarise(
      y = max(.data[[metric_name]], na.rm = TRUE) * 1.10,
      .groups = "drop"
    )

  welch_plot_df <- welch_table %>%
    left_join(y_positions, by = "treatment") %>%
    mutate(
      yend = y,
      label_y = y * 1.04
    )

  y_max <- max(
    welch_plot_df$label_y,
    plot_df[[metric_name]],
    na.rm = TRUE
  )

  plot <- ggplot(
    plot_df,
    aes(
      x = treatment,
      y = .data[[metric_name]],
      fill = genotype,
      colour = genotype
    )
  ) +
    geom_boxplot(
      position = position_dodge(0.7),
      width = 0.45,
      alpha = 0.35,
      linewidth = 0.7,
      outlier.shape = NA
    ) +
    geom_jitter(
      position = position_jitterdodge(
        jitter.width = 0.12,
        dodge.width = 0.7
      ),
      size = 1.6,
      alpha = 0.9
    ) +
    geom_segment(
      data = welch_plot_df,
      aes(x = x, xend = xend, y = y, yend = yend),
      inherit.aes = FALSE,
      linewidth = 0.5,
      colour = "black"
    ) +
    geom_text(
      data = welch_plot_df,
      aes(x = treatment_x, y = label_y, label = p_label),
      inherit.aes = FALSE,
      size = 4
    ) +
    scale_fill_manual(values = setNames(unname(genotype_cols), genotype_labels)) +
    scale_colour_manual(values = setNames(unname(genotype_cols), genotype_labels)) +
    coord_cartesian(ylim = c(0, y_max * 1.15), clip = "off") +
    scamp_theme +
    labs(
      x = "Treatment",
      y = y_label,
      title = plot_title
    )

  list(
    plot = plot,
    anova = anova_table,
    welch = welch_table,
    data = plot_df
  )
}
```

# 5. SCAMP fluorescence metric analyses

```{r fluorescence-metric-analyses}
integrated_result <- make_grouped_metric_analysis(
  data = scamp_metric_df,
  metric = integrated_fluorescence,
  y_label = "Integrated fluorescence (a.u.)",
  plot_title = "Integrated fluorescence",
  metric_id = "integrated_fluorescence"
)

positive_area_result <- make_grouped_metric_analysis(
  data = scamp_metric_df,
  metric = positive_area,
  y_label = paste0(
    "Positive area (%)\nThreshold = ",
    round(positive_threshold, 2),
    " a.u."
  ),
  plot_title = "Extent of spinal mineralization",
  metric_id = "positive_area"
)

mean_positive_result <- make_grouped_metric_analysis(
  data = scamp_metric_df,
  metric = mean_positive_fluorescence,
  y_label = paste0(
    "Mean positive fluorescence (a.u.)\nThreshold = ",
    round(positive_threshold, 2),
    " a.u."
  ),
  plot_title = "Mineralization density",
  metric_id = "mean_positive_fluorescence"
)

p_integrated <- integrated_result$plot
p_positive_area <- positive_area_result$plot
p_density <- mean_positive_result$plot

fluorescence_anova_results <- bind_rows(
  integrated_result$anova,
  positive_area_result$anova,
  mean_positive_result$anova
)

fluorescence_welch_results <- bind_rows(
  integrated_result$welch,
  positive_area_result$welch,
  mean_positive_result$welch
)

write.csv2(
  fluorescence_anova_results,
  file.path(statistics_dir, "SCAMP_fluorescence_two_way_ANOVA_results.csv"),
  row.names = FALSE
)

write.csv2(
  fluorescence_welch_results,
  file.path(statistics_dir, "SCAMP_fluorescence_within_treatment_Welch_BH_results.csv"),
  row.names = FALSE
)

kable(fluorescence_anova_results, caption = "SCAMP fluorescence metric two-way ANOVA results")
kable(fluorescence_welch_results, caption = "SCAMP fluorescence metric within-treatment Welch tests with BH correction")
```

# 6. Integrated fluorescence variability analysis

```{r integrated-levene}
## Brown-Forsythe Levene test: genotype-specific variance comparison
## within each treatment for integrated fluorescence.
## This tests whether inter-individual variability differs between genotypes.

integrated_levene_results <- scamp_metric_df %>%
  group_by(treatment) %>%
  summarise(
    p_raw = safe_levene_p(
      cur_data(),
      value_col = "integrated_fluorescence",
      group_col = "genotype"
    ),
    .groups = "drop"
  ) %>%
  mutate(
    metric_id = "integrated_fluorescence",
    metric = "Integrated fluorescence",
    test = "Brown-Forsythe Levene test; center = median",
    p_adj_BH = p.adjust(p_raw, method = "BH"),
    p_label = format_p_stars(p_adj_BH),
    p_raw_formatted = format_p_value(p_raw),
    p_adj_BH_formatted = format_p_value(p_adj_BH)
  ) %>%
  select(metric_id, metric, treatment, test, p_raw, p_adj_BH, p_label, everything())

write.csv2(
  integrated_levene_results,
  file.path(statistics_dir, "SCAMP_integrated_fluorescence_Levene_BrownForsythe_BH_results.csv"),
  row.names = FALSE
)

kable(integrated_levene_results, caption = "Integrated fluorescence variance comparison between genotypes within treatment")
```

# 7. Ribbon plot of spatial mineralization pattern

```{r ribbon-plot, fig.width=12, fig.height=5}
long_df <- scamp_metric_df %>%
  select(fish, genotype, treatment) %>%
  bind_cols(as.data.frame(pixel_data)) %>%
  pivot_longer(
    cols = starts_with("V"),
    names_to = "Pixel",
    values_to = "Intensity"
  ) %>%
  mutate(
    Pixel = as.numeric(gsub("V", "", Pixel))
  )

ribbon_df <- long_df %>%
  group_by(Pixel, genotype, treatment) %>%
  summarise(
    n = sum(!is.na(Intensity)),
    mean = mean(Intensity, na.rm = TRUE),
    sd = sd(Intensity, na.rm = TRUE),
    sem = sd / sqrt(n),
    CI_lower = ifelse(
      n > 1,
      pmax(mean - qt(0.975, n - 1) * sem, 0),
      NA_real_
    ),
    CI_upper = ifelse(
      n > 1,
      mean + qt(0.975, n - 1) * sem,
      NA_real_
    ),
    .groups = "drop"
  )

positive_regions <- ribbon_df %>%
  group_by(genotype, treatment) %>%
  arrange(Pixel, .by_group = TRUE) %>%
  mutate(
    positive = mean > positive_threshold,
    region = cumsum(positive != lag(positive, default = FALSE))
  ) %>%
  filter(positive) %>%
  group_by(genotype, treatment, region) %>%
  summarise(
    xmin = min(Pixel),
    xmax = max(Pixel),
    .groups = "drop"
  )

p_ribbon <- ggplot(
  ribbon_df,
  aes(
    x = Pixel,
    y = mean,
    colour = genotype,
    fill = genotype
  )
) +
  geom_rect(
    data = positive_regions,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = genotype),
    inherit.aes = FALSE,
    alpha = 0.08,
    colour = NA
  ) +
  geom_ribbon(
    aes(ymin = CI_lower, ymax = CI_upper),
    alpha = 0.30,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  geom_hline(
    yintercept = positive_threshold,
    linetype = "dashed",
    colour = "grey35"
  ) +
  facet_wrap(~ treatment, nrow = 2) +
  scale_colour_manual(values = setNames(unname(genotype_cols), genotype_labels)) +
  scale_fill_manual(values = setNames(unname(genotype_cols), genotype_labels)) +
  coord_cartesian(ylim = c(0, max(ribbon_df$CI_upper, na.rm = TRUE) * 1.05)) +
  scamp_theme +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  ) +
  labs(
    x = "Normalised spine position",
    y = "Mean fluorescence intensity (a.u.)",
    title = "Spatial mineralization profile"
  )

p_ribbon
```

# 8. Calcified vertebrae analysis

```{r calcified-vertebrae-analysis}
if (!file.exists(calcified_data_path)) {
  warning("Calcified vertebrae input file not found: ", calcified_data_path)
  calc_df <- NULL
} else {
  calc_raw <- read.csv2(calcified_data_path, header = TRUE, check.names = TRUE)

  calc_df <- calc_raw %>%
    mutate(
      genotype = factor(
        Genotype,
        levels = names(genotype_labels),
        labels = genotype_labels
      ),
      treatment = factor(
        Treatment,
        levels = treatment_levels,
        labels = treatment_labels
      ),
      calcified_vertebrae = as.numeric(`Calcified.vertebrae.all`)
    ) %>%
    filter(!is.na(genotype), !is.na(treatment), !is.na(calcified_vertebrae))

  calcified_result <- make_grouped_metric_analysis(
    data = calc_df,
    metric = calcified_vertebrae,
    y_label = "Calcified vertebrae all (number)",
    plot_title = "Calcified vertebrae count",
    metric_id = "calcified_vertebrae_all"
  )

  p_calcified_vertebrae <- calcified_result$plot

  write.csv2(
    calc_df,
    file.path(statistics_dir, "SCAMP_calcified_vertebrae_per_fish.csv"),
    row.names = FALSE
  )

  write.csv2(
    calcified_result$anova,
    file.path(statistics_dir, "SCAMP_calcified_vertebrae_two_way_ANOVA_results.csv"),
    row.names = FALSE
  )

  write.csv2(
    calcified_result$welch,
    file.path(statistics_dir, "SCAMP_calcified_vertebrae_within_treatment_Welch_BH_results.csv"),
    row.names = FALSE
  )

  kable(calcified_result$anova, caption = "Calcified vertebrae two-way ANOVA results")
  kable(calcified_result$welch, caption = "Calcified vertebrae within-treatment Welch tests with BH correction")


  p_calcified_vertebrae
}
```

# 9. Supplementary treatment-response analyses within genotype

This section generates SI-style plots where each calcification-related metric is faceted by genotype. Within each genotype, treatment groups are compared against the baseline condition (`-MB -PPi`) using Welch's t-tests followed by Benjamini-Hochberg correction.

```{r si-baseline-treatment-functions}
## -----------------------------
## SI helper: treatment comparisons against baseline within genotype
## -----------------------------

treatment_cols <- c(
  "-MB\n-PPi" = "#4D4D4D",
  "+MB\n-PPi" = "#377EB8",
  "-MB\n+PPi" = "#4DAF4A",
  "+MB\n+PPi" = "#E69F00"
)

safe_welch_baseline_p <- function(data,
                                  value_col,
                                  group_col = "treatment",
                                  baseline_level,
                                  comparison_level) {

  ## Keep only the baseline and the current comparison group, then rebuild
  ## a two-level factor. This is important because treatment is usually stored
  ## as a factor with all treatment levels; t.test() requires exactly two
  ## observed groups and can otherwise fail or return NA through tryCatch().
  local_df <- data %>%
    filter(
      as.character(.data[[group_col]]) %in% c(baseline_level, comparison_level),
      !is.na(.data[[value_col]])
    ) %>%
    mutate(
      comparison_group = factor(
        as.character(.data[[group_col]]),
        levels = c(baseline_level, comparison_level)
      )
    ) %>%
    droplevels()

  if (nlevels(local_df$comparison_group) < 2) return(NA_real_)

  split_x <- split(
    local_df[[value_col]],
    local_df$comparison_group,
    drop = TRUE
  )

  if (length(split_x) < 2) return(NA_real_)

  if (any(sapply(split_x, function(v) sum(!is.na(v)) < 2))) return(NA_real_)

  if (sd(local_df[[value_col]], na.rm = TRUE) < .Machine$double.eps) return(NA_real_)

  tryCatch(
    t.test(
      local_df[[value_col]] ~ local_df$comparison_group,
      var.equal = FALSE
    )$p.value,
    error = function(e) NA_real_
  )
}

make_baseline_treatment_analysis <- function(data,
                                             metric,
                                             y_label,
                                             plot_title,
                                             metric_id,
                                             baseline_label = treatment_labels[1]) {

  metric <- rlang::ensym(metric)
  metric_name <- rlang::as_string(metric)

  plot_df <- data %>%
    filter(
      !is.na(.data[[metric_name]]),
      !is.na(genotype),
      !is.na(treatment)
    ) %>%
    mutate(
      treatment = factor(treatment, levels = treatment_labels),
      genotype = factor(genotype, levels = genotype_labels)
    )

  comparison_levels <- setdiff(treatment_labels, baseline_label)

  stat_df <- expand.grid(
    genotype = levels(plot_df$genotype),
    comparison = comparison_levels,
    stringsAsFactors = FALSE
  ) %>%
    as_tibble() %>%
    rowwise() %>%
    mutate(
      p_raw = {
        current_genotype <- genotype

        safe_welch_baseline_p(
          data = plot_df %>% filter(genotype == current_genotype),
          value_col = metric_name,
          group_col = "treatment",
          baseline_level = baseline_label,
          comparison_level = comparison
        )
      },
      n_baseline = {
        current_genotype <- genotype

        sum(
          plot_df$genotype == current_genotype &
            plot_df$treatment == baseline_label &
            !is.na(plot_df[[metric_name]])
        )
      },
      n_comparison = {
        current_genotype <- genotype

        sum(
          plot_df$genotype == current_genotype &
            plot_df$treatment == comparison &
            !is.na(plot_df[[metric_name]])
        )
      }
    ) %>%
    ungroup() %>%
    group_by(genotype) %>%
    mutate(
      p_adj_BH = p.adjust(p_raw, method = "BH"),
      p_label = format_p_stars(p_adj_BH),
      p_raw_formatted = format_p_value(p_raw),
      p_adj_BH_formatted = format_p_value(p_adj_BH),
      comparison_rank = row_number(),
      x = match(baseline_label, treatment_labels),
      xend = match(comparison, treatment_labels),
      label_x = (x + xend) / 2
    ) %>%
    ungroup() %>%
    mutate(
      metric_id = metric_id,
      metric = plot_title,
      baseline = baseline_label
    )

  y_positions <- plot_df %>%
    group_by(genotype) %>%
    summarise(
      y_base = max(.data[[metric_name]], na.rm = TRUE),
      .groups = "drop"
    )

  stat_plot_df <- stat_df %>%
    left_join(y_positions, by = "genotype") %>%
    group_by(genotype) %>%
    arrange(xend, .by_group = TRUE) %>%
    mutate(
      y = y_base * (1.10 + 0.10 * (row_number() - 1)),
      yend = y,
      label_y = y * 1.035
    ) %>%
    ungroup()

  y_max <- max(
    stat_plot_df$label_y,
    plot_df[[metric_name]],
    na.rm = TRUE
  )

  plot <- ggplot(
    plot_df,
    aes(
      x = treatment,
      y = .data[[metric_name]],
      fill = treatment,
      colour = treatment
    )
  ) +
    geom_boxplot(
      width = 0.55,
      alpha = 0.35,
      linewidth = 0.7,
      outlier.shape = NA
    ) +
    geom_jitter(
      width = 0.12,
      size = 1.6,
      alpha = 0.9
    ) +
    geom_segment(
      data = stat_plot_df,
      aes(x = x, xend = xend, y = y, yend = yend),
      inherit.aes = FALSE,
      linewidth = 0.5,
      colour = "black"
    ) +
    geom_text(
      data = stat_plot_df,
      aes(x = label_x, y = label_y, label = p_label),
      inherit.aes = FALSE,
      size = 3.6
    ) +
    facet_wrap(~genotype, nrow = 1) +
    scale_fill_manual(values = treatment_cols) +
    scale_colour_manual(values = treatment_cols) +
    coord_cartesian(ylim = c(0, y_max * 1.12), clip = "off") +
    scamp_theme +
    labs(
      x = "Treatment",
      y = y_label,
      title = plot_title,
      subtitle = paste0("Within-genotype comparison against baseline: ", gsub("\\n", " ", baseline_label))
    ) +
    theme(
      legend.position = "none",
      strip.text = element_text(face = "bold"),
      axis.text.x = element_text(angle = 0, hjust = 0.5)
    )

  list(
    plot = plot,
    statistics = stat_df %>%
      select(
        metric_id, metric, genotype, baseline, comparison,
        n_baseline, n_comparison,
        p_raw, p_adj_BH, p_label,
        p_raw_formatted, p_adj_BH_formatted
      ),
    data = plot_df
  )
}
```

```{r si-baseline-treatment-analyses, fig.width=12, fig.height=10}
## -----------------------------
## SCAMP fluorescence metrics: treatment effects within each genotype
## -----------------------------

si_integrated_result <- make_baseline_treatment_analysis(
  data = scamp_metric_df,
  metric = integrated_fluorescence,
  y_label = "Integrated fluorescence (a.u.)",
  plot_title = "Integrated fluorescence",
  metric_id = "integrated_fluorescence"
)

si_positive_area_result <- make_baseline_treatment_analysis(
  data = scamp_metric_df,
  metric = positive_area,
  y_label = paste0(
    "Positive area (%)\nThreshold = ",
    round(positive_threshold, 2),
    " a.u."
  ),
  plot_title = "Positive area",
  metric_id = "positive_area"
)

si_density_result <- make_baseline_treatment_analysis(
  data = scamp_metric_df,
  metric = mean_positive_fluorescence,
  y_label = paste0(
    "Mean positive fluorescence (a.u.)\nThreshold = ",
    round(positive_threshold, 2),
    " a.u."
  ),
  plot_title = "Mean positive fluorescence",
  metric_id = "mean_positive_fluorescence"
)

si_baseline_stats <- bind_rows(
  si_integrated_result$statistics,
  si_positive_area_result$statistics,
  si_density_result$statistics
)

if (exists("calc_df") && !is.null(calc_df)) {
  si_calcified_result <- make_baseline_treatment_analysis(
    data = calc_df,
    metric = calcified_vertebrae,
    y_label = "Calcified vertebrae all (number)",
    plot_title = "Calcified vertebrae count",
    metric_id = "calcified_vertebrae_all"
  )

  si_baseline_stats <- bind_rows(
    si_baseline_stats,
    si_calcified_result$statistics
  )
}

write.csv2(
  si_baseline_stats,
  file.path(statistics_dir, "SI_within_genotype_treatment_vs_baseline_Welch_BH_results.csv"),
  row.names = FALSE
)

write.csv2(
  si_baseline_stats %>% filter(metric_id != "calcified_vertebrae_all"),
  file.path(statistics_dir, "SI_SCAMP_metrics_within_genotype_treatment_vs_baseline_Welch_BH_results.csv"),
  row.names = FALSE
)

if (exists("si_calcified_result")) {
  write.csv2(
    si_calcified_result$statistics,
    file.path(statistics_dir, "SI_calcified_vertebrae_within_genotype_treatment_vs_baseline_Welch_BH_results.csv"),
    row.names = FALSE
  )
}

p_si_integrated <- si_integrated_result$plot
p_si_positive_area <- si_positive_area_result$plot
p_si_density <- si_density_result$plot
if (exists("si_calcified_result")) {
  p_si_calcified_vertebrae <- si_calcified_result$plot
}

si_calcification_metrics_plot <- if (exists("p_si_calcified_vertebrae")) {
  (p_si_integrated / p_si_positive_area / p_si_density / p_si_calcified_vertebrae) +
    plot_annotation(tag_levels = "A")
} else {
  (p_si_integrated / p_si_positive_area / p_si_density) +
    plot_annotation(tag_levels = "A")
}

si_calcification_metrics_plot
```

```{r save-si-baseline-treatment-figures}
ggsave(
  file.path(figure_dir, "SI_calcification_metrics_by_genotype_treatment_vs_baseline.pdf"),
  si_calcification_metrics_plot,
  width = 12,
  height = if (exists("p_si_calcified_vertebrae")) 16 else 12,
  device = cairo_pdf
)

ggsave(
  file.path(figure_dir, "SI_calcification_metrics_by_genotype_treatment_vs_baseline.png"),
  si_calcification_metrics_plot,
  width = 12,
  height = if (exists("p_si_calcified_vertebrae")) 16 else 12,
  dpi = 600
)
```

# 10. Biomarker plots

```{r biomarker-plots, fig.width=12, fig.height=8}
p_integrated
p_positive_area
p_density
if (exists("p_calcified_vertebrae")) p_calcified_vertebrae
```

# 11. Combined statistics export

```{r combined-statistics-export}
if (exists("calcified_result")) {
  combined_anova_results <- bind_rows(
    fluorescence_anova_results,
    calcified_result$anova
  )

  combined_welch_results <- bind_rows(
    fluorescence_welch_results,
    calcified_result$welch
  )
} else {
  combined_anova_results <- fluorescence_anova_results
  combined_welch_results <- fluorescence_welch_results
}

write.csv2(
  combined_anova_results,
  file.path(statistics_dir, "SCAMP_all_two_way_ANOVA_results.csv"),
  row.names = FALSE
)

write.csv2(
  combined_welch_results,
  file.path(statistics_dir, "SCAMP_all_within_treatment_Welch_BH_results.csv"),
  row.names = FALSE
)

kable(combined_anova_results, caption = "Combined two-way ANOVA results")
kable(combined_welch_results, caption = "Combined within-treatment Welch tests with BH correction")
```

# 12. Final figure export

```{r final-export, fig.width=12, fig.height=10}
if (exists("p_calcified_vertebrae")) {
  final_figure <-
    p_ribbon /
    ((p_integrated | p_positive_area) /
       (p_density | p_calcified_vertebrae)) +
    plot_layout(heights = c(1.1, 1.6)) +
    plot_annotation(
      title = "SCAMP mineralization analysis",
      tag_levels = "A"
    )
} else {
  final_figure <-
    p_ribbon /
    (p_integrated | p_positive_area | p_density) +
    plot_layout(heights = c(1.2, 1)) +
    plot_annotation(
      title = "SCAMP mineralization analysis",
      tag_levels = "A"
    )
}

ggsave(
  filename = file.path(figure_dir, "SCAMP_MB_PPi_summary_figure.pdf"),
  plot = final_figure,
  width = 12,
  height = 10,
  units = "in",
  device = cairo_pdf
)

ggsave(
  filename = file.path(figure_dir, "SCAMP_MB_PPi_summary_figure.png"),
  plot = final_figure,
  width = 12,
  height = 10,
  units = "in",
  dpi = 600
)

final_figure
```

# 13. Session information

```{r session-info}
writeLines(
  capture.output(sessionInfo()),
  con = file.path(qc_dir, "sessionInfo.txt")
)

sessionInfo()
```
